Wait-free Queues with Polygarithmic Step Complexity

### 2022-05-25 article/psuedocode.tex
apply may 20 feedback

\framebox[1.1\width]{TEST} As a temporary test I have changed the name of n.size to n.established here, other options are n.head and n.lastBLock but they might be confusing since we have used them before. Fix the logical order of definitions (cyclic refrences).

@@

\framebox[1.1\width]{TEST} How to merge notions of blocks and operations? block b $\in$ block c means b is subblock of c. block b $\in$ set B means b is in B. Merge these two to have shorter formulaes.

@@

\begin{definition}[Subblock]
Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}. Block \texttt{b} is a subblock of a \texttt{n.blocks[i]} if it is a direct subblock of it or subblock of a direct subblock of it.
\end{definition}

@@

\begin{definition}
  Block \nf{b} in \nf{n.blocks} is \emph{Established} at time $t$ if \nf{n.established} is greater than index of \nf{b} at time $t$. Block \nf{b} is in \emph{EST\sub{n, t}} if \nf{b} is a subblock of \nf{b$^\prime$} in \nf{n.blocks} such that \nf{b$^\prime$} is established  at time $t$.
\end{definition}

@@

\begin{lemma}
  Each block is not duplicated.
\end{lemma}
\begin{proof}
  \nf{n.CreateBlock(i)} is successful for \nf{i=1,2,3,...} exactly one time and by definition of \nf{CreateBlock} the range of their blocks are disjoint so in propagation procedure on block can not be propagated twice (reference to propagation).
\end{proof}

\begin{lemma}[headProgress] \label{lem::sizeProgress}
\nf{n.established} is non-decreasing over time.
\end{lemma}
\begin{proof}
  Induced trivially from the pseudocode since \nf{n.established} is only incremented.
\end{proof}

\begin{lemma}[headPosition] \label{lem::headPosition} If the value read in Line \ref{prevLine}(\nf{h=n.established}) is $h$ then, \nf{n.blocks[i]=null} for \nf{i>h} and \nf{n.blocks[i]$\neq$null} for \nf{i<h}.
\end{lemma}
\begin{proof}
At the end of every \nf{n.Refresh()} with a block  with size greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}). If a process went to sleep before incrementing the \nf{established} (line \ref{incrementHead1}), nothing can be appended to \nf{n.blocks} until another process increments \nf{n.establsihed} (Line \ref{incrementHead2}).
\end{proof}

@@

\begin{lemma}[createBlock] \label{lem::createBlock}
  If \nf{n.CreateBlock(i)} is invoked at time $t$ and returns block $new$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(new)}.
\end{lemma}
\begin{proof}
From lemmas 7,8 we know that \nf{n.blocks[i-1]} is full and assume it is invoked at $t^\prime$. 
We want to prove every block $b$ that is $\in$ \nf{EST\sub{n.left,t}} $\cup$ \nf{EST\sub{n.right,t}} is also $\in$ \nf{EST\sub{n,t}}$^\prime$ $\cup$ \nf{new}. It is easy to see \nf{EST\sub{n.left,t}} $\cap$ \nf{EST\sub{n.right,t}} is $\emptyset$. If $b$ is $\in$ \nf{EST\sub{n.left,t$^\prime$}} it is simple to see. Else we can see from the code of the \nf{CreateBlock} that \nf{new} contains it. Right child is the same.
% WLOG we prove the claim for the left child. Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are established at time $t$.  Since the head is only increasing (Lemma~\ref{lem::sizeProgress}) the  claim holds after $t$. See Figure~\ref{fig::createBlock}. The right child is the same.
\end{proof}

\begin{lemma}[trueRefresh] \label{lem::trueRefresh}
  Let $t_i$ be the time \nf{n.Refresh()} is invoked and $t_t$ be the time it is terminated. Suppose \nf{n.Refresh()}'s \nf{TryAppend(new, s)} returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})}.
\end{lemma}
\begin{proof}
Suppose the previous true returning \nf{n.Refresh()} took place in time $(t_i^\prime-t_t^\prime)$. By induction on the number of succsuful \nf{Refresh()}'s \nf{n} we know that \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})}. Operations that are in \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})} but not in \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})} are in the block returned by \nf{n.CreateBlock()} by lemma 10. 
By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readSize} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
\end{proof}

\begin{lemma}[falseRefresh] \label{lem:falseRefresh}
    If instance \nf{r} of \nf{n.Refresh()} reads value \nf{s} on line \ref{readSize} and then returns \nf{false}, then there is another instance $r^\prime$ of \nf{n.Refresh()} that has performed a successful \nf{TryAppend(new, s)}. A \nf{TryAppend()} is successful if its \nf{CAS} is successful.
\end{lemma}
\begin{proof}
  If there is no other concurrent successful \nf{n.Refresh()} then \nf{n.Refresh()} would succeed in Line 313. So there is another \nf{n.Refresh()}, that has appended its block in \nf{n.blocks[h]} after Line 310 of the first \nf{n.Refresh()}. 
%Otherwise the other \nf{Refresh(n)} should have read \nf{h$^\prime$>h} instead of \nf{h} for \nf{n.head}(Line 52).
\end{proof}

@@

If Line 313 of $r_1$ or $r_2$ returns \nf{true}, then the claim is held by Lemma~\ref{lem::trueRefresh}. If not, then there is another successful instance of \nf{n.Refresh()} $r_1^\prime$ by Lemma~\ref{lem:falseRefresh} for $r_2$. From $t_{i_{r_2}^\prime} > t_1$ and Lemma~\ref{lem::trueRefresh}, \ref{lem::establishedOrder} we know that \nf{ops(EST\sub{n.left, t\sub{1}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{1}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{2}})}.
%It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}). If $R_2$ succeeds by Lemma~\ref{lem::trueRefresh} the claim holds. If not, it is deduced that \nf{n.blocks[h]} was not \nf{null} before $R_2$'s \nf{CAS}. Furthermore, \nf{n.blocks[h]} was \nf{null} before reading \nf{h} by $R_1$. So there is a successful \nf{Refresh()} after the read of \nf{h} in $R_1$ and before the \nf{CAS} of $R_2$. This \nf{Refresh()} contains all the new established operations before Line 35 (Maybe Lemma~\ref{lem::oldnewOrder}) and by Lemma~\ref{lem::trueRefresh} our claim holds. See Figure~\ref{}.
\end{proof}

\begin{figure}[hbt]
  \center\includegraphics[width=3in]{pics/doublyrefresh.drawio.png}
  \caption{$R_2^\prime$'s \nf{CAS} is executed after \nf{h1=n.head}.}
\end{figure}

>>>>

\framebox[1.1\width]{TEST} As a temporary test I have changed the name of n.size to n.head here, other options are n.head and n.lastBLock but they might be confusing since we have used them before. Fix the logical order of definitions (cyclic refrences).

@@

\framebox[1.1\width]{TEST} How to merge notions of blocks and operations? block b $\sqsubseteq$ block c means b is subblock of c. block b $\in$ set B means b is in B. Merge these two to have shorter formulaes.

@@

\begin{definition}[Subblock]
Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}. Block \texttt{b} is a subblock of a \texttt{n.blocks[i]} if it is a direct subblock of it or subblock of a direct subblock of it. Block b is direct superblock of block c if c is direct subblock of b.
\end{definition}

@@

\begin{definition}
  Block \nf{b} in \nf{n.blocks} is \emph{Established} at time $t$ if \nf{n.head} is greater than index of \nf{b} at time $t$. Block \nf{b} is in \emph{EST\sub{n, t}} if \nf{b} is a subblock of \nf{b$^\prime$} in \nf{n.blocks} such that \nf{b$^\prime$} is established  at time $t$.
\end{definition}

@@

\begin{lemma}[headProgress] \label{lem::sizeProgress}
\nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left, right}} is non-decreasing over \nf{i}.
\end{lemma}
\begin{proof}
  Induced trivially from the pseudocode since \nf{n.head} is only incremented. \nf{n.blocks[i]. Also end\sub{left, right}} are greater than or equal the previous values in the \nf{CreateBlock(i)} code (Lines~\ref{lastLine, prevLine}).
\end{proof}

\begin{lemma}
Every block has most one direct superblock.
\end{lemma}
\begin{proof}
To show this we are going to refer to the way n.blocks[] is partitioned while propagating up to n.parent. \nf{n.CreateBlock(i)} merges the blocks in \nf{n.left.blocks[n.blocks[i-1].end\sub{left}..n.blocks[i].end\sub{left}]} and \nf{n.right.blocks[n.blocks[i-1].end\sub{right}..n.blocks[i].end\sub{right}]} (Lines \ref{lastLine, pr}). From we know that \nf{end\sub{left},end\sub{right}} are non decreasing, so the range of the subblocks of n.blocks are disjoint.
\end{proof}

\begin{lemma}[headPosition] \label{lem::headPosition} If the value read in Line \ref{prevLine}(\nf{h=n.head}) is $h$ then, \nf{n.blocks[i]=null} for \nf{i>h} and \nf{n.blocks[i]$\neq$null} for \nf{i<h}.
\end{lemma}
\begin{proof}
At the end of every \nf{n.Refresh()} with a block  with size greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}). If a process went to sleep before incrementing the \nf{established} (line \ref{incrementHead1}), nothing can be appended to \nf{n.blocks} until another process increments \nf{n.establsihed} (Line \ref{incrementHead2}).
\end{proof}

@@

%\begin{lemma}[createBlock] \label{lem::createBlock}
%  If \nf{n.CreateBlock(i)} is invoked at time $t$ and returns block $new$, then \nf{ops(EST\sub{n.left, t})} $\cup$ \nf{ops(EST\sub{n.right, t})} - \nf{ops(EST\sub{n, t})} $\subseteq$ \nf{ops(new)}.
%\end{lemma}
%\begin{proof}
%From lemmas 7,8 we know that \nf{n.blocks[i-1]} is full and assume it is invoked at $t^\prime$. 
%We want to prove every block $b$ that is $\in$ \nf{EST\sub{n.left,t}} $\cup$ \nf{EST\sub{n.right,t}} is also $\in$ \nf{EST\sub{n,t}}$^\prime$ $\cup$ \nf{new}. It is easy to see \nf{EST\sub{n.left,t}} $\cap$ \nf{EST\sub{n.right,t}} is $\emptyset$. If $b$ is $\in$ \nf{EST\sub{n.left,t$^\prime$}} it is simple to see. Else we can see from the code of the \nf{CreateBlock} that \nf{new} contains it. Right child is the same.
%% WLOG we prove the claim for the left child. Blocks in \nf{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} are established at time $t$.  Since the head is only increasing (Lemma~\ref{lem::sizeProgress}) the  claim holds after $t$. See Figure~\ref{fig::createBlock}. The right child is the same.
%\end{proof}

\begin{lemma}[trueRefresh] \label{lem::trueRefresh}
  Let $t_i$ be the time \nf{n.Refresh()} is invoked and $t_t$ be the time it is terminated. Suppose \nf{n.Refresh()}'s \nf{TryAppend(new, s)} returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{t}})}.
\end{lemma}
\begin{proof}
From Lemma \ref{lem::establishedOrder} we know that \nf{ops(EST\sub{n, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}. So it remains to show \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} - \nf{ops(EST\sub{n, t\sub{i}})} which we call new operations $\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}. If \nf{TryAppend}returns true a block is appended to n. From the code of the \nf{CreatBlock} this block includes the establsihed blocks in n's children at $t_i$. Since the head in createblock is read after $t_i$. So the new operations are inthe block appended to $n$.

%Suppose the previous true returning \nf{n.Refresh()} took place in time $(t_i^\prime-t_t^\prime)$. By induction on the number of succsuful \nf{Refresh()}'s \nf{n} we know that \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})}. Operations that are in \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})} but not in \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})} are in the block returned by \nf{n.CreateBlock()} by lemma 10. 
%By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readSize} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
\end{proof}

%\begin{lemma}[falseRefresh] \label{lem:falseRefresh}
%    If instance \nf{r} of \nf{n.Refresh()} reads value \nf{s} on line \ref{readSize} and then returns \nf{false}, then there is another instance $r^\prime$ of \nf{n.Refresh()} that has performed a successful \nf{TryAppend(new, s)}. A \nf{TryAppend()} is successful if its \nf{CAS} is successful.
%\end{lemma}
%\begin{proof}
%  If there is no other concurrent successful \nf{n.Refresh()} then \nf{n.Refresh()} would succeed in Line 313. So there is another \nf{n.Refresh()}, that has appended its block in \nf{n.blocks[h]} after Line 310 of the first \nf{n.Refresh()}. 
%%Otherwise the other \nf{Refresh(n)} should have read \nf{h$^\prime$>h} instead of \nf{h} for \nf{n.head}(Line 52).
%\end{proof}

@@

If Line 313 of $r_1$ or $r_2$ returns \nf{true}, then the claim is held by Lemma~\ref{lem::trueRefresh}. If not, then there is another successful instance of \nf{n.Refresh()} $r_2^\prime$ which has did TryAppend successfully into \nf{n.blocks[i+1]}. In figure 1 we see why the block it is appending contains established block in the n's children at $t_i$, since it create a block reading the head after $t_1$.
%It is obvious that the \nf{new} constructed by the second Refresh in Line 36 contains the blocks in \nf{new} by $R_1$ which $R_1^\prime$ did not contain, since \nf{n.head} is only increasing (Maybe Lemma~\ref{lem::oldnewOrder}). If $R_2$ succeeds by Lemma~\ref{lem::trueRefresh} the claim holds. If not, it is deduced that \nf{n.blocks[h]} was not \nf{null} before $R_2$'s \nf{CAS}. Furthermore, \nf{n.blocks[h]} was \nf{null} before reading \nf{h} by $R_1$. So there is a successful \nf{Refresh()} after the read of \nf{h} in $R_1$ and before the \nf{CAS} of $R_2$. This \nf{Refresh()} contains all the new established operations before Line 35 (Maybe Lemma~\ref{lem::oldnewOrder}) and by Lemma~\ref{lem::trueRefresh} our claim holds. See Figure~\ref{}.
\end{proof}

\begin{figure}[hbt]
  \center\includegraphics[width=6in]{pics/completedouble.png}
  \caption{$t_t$ is after $R_2^\prime$'s reads n.head after n.head is incremented which is after the succesufil TryAppend(i) which is after $R_1$'s reading n.head that takes place at $t_i$.}
\end{figure}
